Reassessing the role of theory and machine learning in forecasting civil conflict

### 2020-09-17 paper.Rmd
first edits
mdw

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves justified as easing interpretation. Smoothing is used to produce all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.

@@

B&S advocate the use of theory to guide prediction. But "theory" as used by B&S is an ambiguous and undefined concept.  It is not a procedure.  They actually create a model with ten right-hand side variables that are supposed to capture a complicated repression-dissent dynamic. There is wide-ranging literature on this dynamic that could justify many specifications.  As such, their baseline comparison is an unfortunate standard bearer for strong theory.

>>>>

---
title: "Reassessing the role of theory and machine learning in forecasting civil conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  While their model is not very theoretical, it is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves justified as easing interpretation. Smoothing is used to produce all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined N articles in K journals over the period from 20xx-2020 looking for articles which have a binary dependent variable and use the ROC techonology in the analysis section. We found that XX percent of these articles use the empirical ROC calculations and only 100-XX use a smoothed curve.  This is not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard.  What is surprising is that anyone would choose the smoothed ROC implementation.
INSERT SMALL STUDY HERE WITH DETAILS.

@@

B\&S advocate the use of theory to guide prediction. But "theory" as used by B\&S is an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  A theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.  For the purposes of discussion we call this strong theory.

However, this is not what many mean by theory in the contemporary social science. For many,
a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables. A theory in this sense could be entirely created by the scholar. More typically it is fused together from a meta-analysis of prior work on the topic under consideration.   We call these types of theory weak theory.

They are weak because they do not include two characteristics of strong theory, though they may include one:
1. A theory is separate from the case or cases being examined. It is intented to be general.
2. A theory generates new statements that in principle can be empirically falsified, not proven.

Throwing a bunch of variables that you think are related to a dependent variable and popping then Into a non parametric model is not a theory-driven model.

It is not a procedure.  They actually create a model with ten right-hand side variables that are supposed to capture a complicated repression-dissent dynamic. There is wide-ranging literature on this dynamic that could justify many specifications.  As such, their baseline comparison is an unfortunate standard bearer for strong theory.
